---
title: 'Math 50 Fall 2017, Homework #2'
output: rmarkdown::github_document
By: Aadil Islam
---

__NOTE: For your homework download and use the template__ (https://math.dartmouth.edu/~m50f17/HW2.Rmd)

__Read the green comments in the rmd file to see where your answers should go.__

## Question-1 (Sample)
- Given a fixed confidence interval percentage (say 95%) at what value of x does CI on the mean response achieve its minimum width?  
- The width of the interval is $2t_{\alpha/2,n-2}\sqrt{MS_{Res}((1/n)+(x_0-\bar{x})^2/S_{xx}}$
- and all terms inside the square root are positive. Therefore it is minimized when $x_0=\bar{x}$. 
- Write an R-chunk using the propellant data which computes the following. 
  - (a) Fit a simple linear regression model relating shear strength to age.
  - (b) Plot scatter diagram.
  - (c) Plot two lines (in blue color) that traces upper and lower limits of 95% confidence interval on $E(y|x_0)$
  - (d) Plot two lines (in red color) that traces upper and lower limits of 95% prediction interval for $y$ 
  - (e) Print the 95% quantile of the corresponding t distribution 

```{r}
prop<-read.table("https://math.dartmouth.edu/~m50f17/propellant.csv", header=T, sep=",")

shearS<-prop$ShearS 
age<-prop$Age

plot(age, shearS, xlab = "Propellant Age (weeks)", ylab = "Shear S. (psi)", main = "Rocket Propellant")
fitted <- lm(shearS ~ age)

ageList <- seq(0,25,0.5)
cList <- predict(fitted, list(age = ageList), int = "c", level = 0.95)
pList <- predict(fitted, list(age = ageList), int = "p", level = 0.95)

matlines(ageList, pList, lty='solid' ,  col = "red")
matlines(ageList, cList, lty = 'solid', col = "blue")

# since n=20 we look at the t_18 distribution
wantedQuantile <- qt( 0.95, 18) ; 
```

95% quantile is of $t_18$ is : `r wantedQuantile` 

## Question-2

- Plot the same graph as in Question-1 without using R function predict, but instead directly calculating the interval limits we discussed in class. 
- In particular, what are the limits of 95% confidence interval on $E(y|x_0)$?  

```{r}
# Computation part of the answer : 

#creating scatter plot
prop<-read.table("https://math.dartmouth.edu/~m50f17/propellant.csv", header=T, sep=",")
shearS<-prop$ShearS 
age<-prop$Age
plot(age, shearS, xlab = "Propellant Age (weeks)", ylab = "Shear S. (psi)", main = "Rocket Propellant")

#preparing to create intervals
fitted <- lm(shearS ~ age)
yhat <- fitted$fitted.values
Sxx <- sum((age-mean(age))^2)
Sxy <- sum(shearS*(age-mean(age)))
beta_1_hat <- Sxy/Sxx
beta_0_hat <- mean(shearS) - beta_1_hat*(mean(age))
ageList <- seq(0,25,0.5)
mean_resp_hat = beta_0_hat + beta_1_hat*ageList
SSres = sum((shearS-yhat)^2)
MSres = SSres/18

#constructing confidence interval
conf_upper = mean_resp_hat + qt(1-0.05/2,18)*(MSres*(1/20 + ((ageList-mean(age))^2/Sxx)))^0.5
conf_lower = mean_resp_hat - qt(1-0.05/2,18)*(MSres*(1/20 + ((ageList-mean(age))^2/Sxx)))^0.5
matlines(ageList, conf_upper, lty='solid' ,  col = "blue")
matlines(ageList, conf_lower, lty = 'solid', col = "blue")

#constructing prediction interval
yhat = beta_0_hat + beta_1_hat*ageList
pred_upper = yhat + qt(1-0.05/2,18)*(MSres*(1+1/20 + ((ageList-mean(age))^2/Sxx)))^0.5
pred_lower = yhat - qt(1-0.05/2,18)*(MSres*(1+1/20 + ((ageList-mean(age))^2/Sxx)))^0.5
matlines(ageList, yhat, lty='solid' ,  col = "purple")
matlines(ageList, pred_upper, lty='solid' ,  col = "red")
matlines(ageList, pred_lower, lty = 'solid', col = "red")

#determining E(y|x0) at x0 = xbar
ageList = mean(age)
mean_resp_hat = beta_0_hat + beta_1_hat*ageList
conf_upper_x_bar = mean_resp_hat + qt(1-0.05/2,18)*(MSres*(1/20 + ((ageList-mean(age))^2/Sxx)))^0.5
conf_lower_x_bar = mean_resp_hat - qt(1-0.05/2,18)*(MSres*(1/20 + ((ageList-mean(age))^2/Sxx)))^0.5

```

Limits of 95% confidence interval on $E(y|x_0)$: (`r conf_upper_x_bar`, `r conf_lower_x_bar`)

## Question-3
- Load the propellant data and fit a simple linear regression model relating shear strength to age.

### Part (a) 
- Test the hypothesis $\beta_1 = -30$ using confidence level 97.5%.

```{r}
#fitting model to data
prop<-read.table("https://math.dartmouth.edu/~m50f17/propellant.csv", header=T, sep=",")
shearS<-prop$ShearS 
age<-prop$Age
plot(age, shearS, xlab = "Propellant Age (weeks)", ylab = "Shear S. (psi)", main = "Rocket Propellant")
fitted <- lm(shearS ~ age)
yhat <- fitted$fitted.values
yBar = mean(shearS)
abline(fitted)

#calculations
Sxx <- sum((age-mean(age))^2)
Sxy <- sum(shearS*(age-mean(age)))
beta_1_hat <- Sxy/Sxx
beta_0_hat <- mean(shearS) - beta_1_hat*(mean(age))
SSres = sum((shearS-yhat)^2)
MSres = SSres/18
SSt <- sum((shearS-yBar)^2)
SE_b1=(MSres/Sxx)^.5
SE_b0=(MSres*(1/20+((mean(age))^2)/Sxx))^.5
```

beta_1 -30 does not lie in the 97.5% confidence interval for $\beta_1$: (-44.22, -30.09). Thus we reject this hypothesis and conclude that $\beta_1 \neq -30$.

### Part (b) 
- Calculate the limits of 97.5% confidence interval for $\beta_0$ and $\beta_1$

```{r}
upper_beta_1 = beta_1_hat + qt(1-0.025/2,df=18)*SE_b1
lower_beta_1 = beta_1_hat - qt(1-0.025/2,df=18)*SE_b1
```

Upper Limit Beta 1: `r upper_beta_1`

Lower Limit Beta 1: `r lower_beta_1`

```{r}
upper_beta_0 = beta_0_hat + qt(1-0.025/2,df=18)*SE_b0
lower_beta_0 = beta_0_hat - qt(1-0.025/2,df=18)*SE_b0
```

Upper Limit Beta 0: `r upper_beta_0`

Lower Limit Beta 0: `r lower_beta_0`

### Part (c) 
- Is there any relation between the answers you find in part (a) and (b) ?

If $\beta_1 = G$ lies within the limits of the confidence interval, one can conclude (with confidence at the given level) that G can be the true $beta_1$. Contrarily, if $\beta_1 = G$ lies outside of the limits of the confidence interval, at the given confidence level we conclude that G cannot be the true $beta_1$. Another method of testing the hypothesis would be via the calculation of p-values, but we should still get the same results as we did with confidence intervals.

### Part (d) 
- Calculate $R^2$

```{r}
Rsq = 1 - SSres/SSt
```

R-squared: `r Rsq`

## Question-4
- Load the propellant data. This time let us consider a relation between square of shear strength and propellant age. 

### Part (a) 
- Fit a simple linear regression model relating __square__ of shear strength to age. Plot scatter diagram and fitted line. 

```{r}
prop<-read.table("https://math.dartmouth.edu/~m50f17/propellant.csv", header=T, sep=",")

shearS_sq<-(prop$ShearS)^2
age<-prop$Age
plot(age, shearS_sq, xlab = "Propellant Age (weeks)", ylab = "Shear S. Squared (psi^2)", main = "Rocket Propellant")
fitted <- lm(shearS_sq ~ age)
abline(fitted,lwd=2,col='blue')
```

### Part (b) 
- Using analysis-of-variance test for significance of regression (using the formulas we discussed in class)

```{r}
yhat <- fitted$fitted.values
SSr <- sum((yhat-mean(shearS_sq))^2)
SSres = sum((shearS_sq-yhat)^2)
MSr = SSr
MSres = SSres / 18
f_stat = MSr/MSres
limit = qf(0.95, 1, 18)
```

We calculate the F-statistic: `r f_stat`, which does not belong in confidence interval qf(0.95, 1, 18): (0, 4.41). Therefore, this is a statistically significant result at 5% level and we reject the null hypothesis of $beta_1 = 0$. Thus, the test supports a linear relationship.

### Part (c) 
- Use t-test and check significance of regression (using the formulas we discussed in class)

```{r}
Sxx <- sum((age-mean(age))^2)
Sxy <- sum(shearS_sq*(age-mean(age)))
beta_1_hat <- Sxy/Sxx
t_stat = abs((beta_1_hat-0)/(MSres/Sxx)^0.5)
limit = qt(1-0.05/2,18)
```

t-statistic: `r t_stat`, while qt(0.975): `r limit`. The t-statistic does not belong in the confidence interval (-2.10, 2.10) at the 5% level, thus we reject the null hypothesis of $beta_1 = 0$. Again, the test supports a linear relationship.

### Part (d) 
- Does the regression analysis predict a linear relationship between square of shear strength and propellant age ?

Yes, according to our results from b) and c), the regression analysis predicts a linear relationship between square of shear strength and propellant age.

## Question-5
- Once again using propellant data fit a simple linear regression model between shear strength and propellant age. 
- Consider the t-test for hypothesis $\beta_1 = G_1$, and develop a test for $\beta_1 > G_1$ instead. Then 

### Part (a)  
- Test the hypothesis $\beta_1 > -50$ with confidence level 99.9%. 

```{r}
prop<-read.table("https://math.dartmouth.edu/~m50f17/propellant.csv", header=T, sep=",")
shearS<-prop$ShearS
age<-prop$Age
plot(age, shearS, xlab = "Propellant Age (weeks)", ylab = "Shear S. (psi)", main = "Rocket Propellant")
fitted <- lm(shearS ~ age)
abline(fitted,lwd=2,col='blue')

Sxx <- sum((age-mean(age))^2)
Sxy <- sum(shearS*(age-mean(age)))
beta_1_hat <- Sxy/Sxx
yhat <- fitted$fitted.values
SSres = sum((shearS-yhat)^2)
MSres = SSres / 18
t_stat = abs((beta_1_hat-(-50))/(MSres/Sxx)^0.5)
p_val = pt(t_stat, 18)
```

It was found that the t-statistic: `r t_stat`. It follows that the p-value: `r p_val`. This is greater than the significance level of $\alpha = 0.001$ and thus we fail to reject the null hypothesis. We conclude at the 0.1% level that the true $beta_1$ is greater than -50.

### Part (b)  
- Find the smallest value $G_1$ such that the hypothesis $\beta_1 > G_1$ is rejected. 

```{r}
guess = -(qt(0.001, 18)*(MSres/Sxx)^0.5-beta_1_hat)
```

G_1: `r guess`

### Part (c)  Similarly what is the smallest value $G_0$ such that the hypothesis $\beta_0 > G_0$ is rejected. 

```{r}
guess = -(qt(0.001, 18)*(MSres*(1/20 + ((mean(age))^2/Sxx)))^0.5-beta_0_hat)
```
